Retrofitting Effect Handlers onto OCaml

### 2020-11-06 retro-concurrency.tex
No perform benchmarks

\newcommand{\rot}[1]{\adjustbox{angle=90}{\textbf{#1}}}

@@

When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
enforce at-most once semantics. Resuming a continuation involves traversing the
linked-list of fibers and making the last fiber point to the current fiber.
Just as in the operational semantics, the implementation invokes the
appropriate closure to either |continue| or |discontinue| the continuation
(position |p3|). We perform standard tail-call optimization so that resumptions
at tail positions do not build up stack.

@@

\subsection{No effects benchmaks}

@@

\begin{table}
\caption{Micro benchmarks without effects. Each entry shows percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-5mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
}
\label{tab:micro}
\end{table}

We also present micro benchmarks results in Table~\ref{tab:micro}. Since micro
benchmarks magnify micro-architectural optimisations, we also report the number
of instructions executed along with time. |exnval| perform 100 million
iterations of installing exception handlers and returning with value.
|exnraise| is similar, but raises an exception in each iteration. |extcall| and
|callback| perform 100 million external calls and callbacks to identity
functions. The other micro benchmarks are highly recursive programs and were
taken from ~\ref{Folklore}. For micro benchmarks, we observed that padding
tightloops with a few |nop| instructions, which changes the loop alignment,
makes the code upto 15\% faster. Hence, the difference in running times under
15\% may not be statistically significant.

@@

>>>>

\newcommand{\rot}[2][75]{\adjustbox{angle=#1}{\textbf{#2}}}

@@

When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
enforce at-most once semantics. Resuming a continuation involves traversing the
linked-list of fibers and making the last fiber point to the current fiber.
Just as in the operational semantics, the implementation invokes the
appropriate closure to either |continue| or |discontinue| the continuation
(position |p3|). We perform standard tail-call optimisation so that resumptions
at tail positions do not build up stack.

@@

\subsection{No effects benchmarks}

@@

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-5mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
\vspace{-3mm}
}
\label{tab:micro_noeffect}
\end{table}

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| perform 100 million iterations of installing exception handlers
and returning with value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from ~\ref{Folklore}. For micro benchmarks,
we observed that padding tightloops with a few |nop| instructions, which
changes the loop alignment, makes the code upto 15\% faster. Hence, the
difference in running times under 15\% may not be statistically significant.

@@

\subsection{No perform benchmarks}

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 10.57 & 15.76 & 16.96 & 12.88 & 13.64 \\
	\texttt{monad} 	& 42.79 & 69.77 & 348.69 & 33.29 & 39.24 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
\end{table}

Next we look at the cost of setting up an effect handler, but not |perform|ing
an effect. The aim is to quantify the cost of setting up and tearing down
effect handlers compared to a non-tail function call. To this end, we surround
the non-tail calls in the recursive micro benchmarks with an effect handler. We
also include a version of the same benchmarks implemented using concurrency
monad~\cite{} (|monad|) as a proxy for CPS versions. Recall that OCaml compiler
does not use CPS in its IR. In the |monad| version, we use a |fork| to invoke
the non-tail call and use an |MVar| to collect its result.

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 13.77$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. Concurrency
monad suffers due to the heap allocation of continuation frames (which need to
be GCed), where as effect handlers benefit from stack allocation of the frames.
We note that a compiler that uses CPS IR will be faster than the concurrency
monad implementation due to optimisations to reduce the heap allocation of
continuation frames. But Farvardin et al.~\cite{} show that CPS with
optimisations is still slower than using the call stack.

\subsection{Concurrent benchmarks}
